---
title: "Homework 1"
author: "Raul Torres Aragon"
date: "2023-01-10"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
load("hw1_objects.RData")
```

## Problem 1  

Investigate the impact of correlation (among observations) on the performance of linear 
regression. Specifically, conduct a series of simulation experiments in which there are $m$ 
individuals who each have $n$ observations.  
For simplicity, you may make the outcomes normally distributed with common correlation $\rho$.  

Simulation set up:  
The idea is to create observation correlated data such that a linear mixed model with a 
random intercept would capture the true $\boldsymbol{beta}$ vector of parameters. Thus I 
first seek to create random effect $b$ parameter (intercept) such that 
$b \sim \mathcal{N}(0, \theta)$ where $\theta = -(\rho \sigma^2)/(\rho - 1)$ and $\sigma=1$.  

I then use `rnorm(0, theta)` R function to generate a $\boldsymbol{b}$ of length $n$, but 
with $m$ distinct values (one per $m$ cluster).  

I then create the random disturbance vector $e$ as $e \sim \mathcal{N}(0, \sigma^2)$.  

Next, I create a "trt" variable where I randomly assign $X=1$ to some clusters and $X=0$ to 
the rest. I then set true relationship between X and Y--with a random effect--as 
$Y_i=\beta_0 + b_i + \beta_1 X_i + e_i$.  

After that, I regress Y on X using R function `lm`, but purspoley not seeking to account 
for the $b_i$ effect. I store the $\hat{\beta}_0$ amd $\hat{\beta}_1$ coefficients, their 
pvalue, and confidence interval to determine significance at $\alpha=0.05$ coverage, and in 
the case of $\beta=0$, type I error probability (rejecting a true null hypothesis).  

I do this 1,000 times per combination of $rho$, $m$, $n$, and where I set $\beta_1=2.3$ and 
$\beta_0 = 0$ for some iterations and $\beta_0=1.9$ for others. I store the above-mentioned 
values in a vectors of length 1,000 (one per iteration), and then take the mean 
of each vector to compute coverage, type I error (when the null is true), and bias of both 
$\hat{\beta}_0$ and $\hat{\beta}_1$ which I do by computing the mean of each 
$\hat{\boldsymbol{\beta}}$ and subtracting $\boldsymbol{\beta}$, the true vector of parameters. 

The following tables show the results for the null model, i.e., $\beta_0=0$. ($\beta_1 = 2.3$.) 

```{r, echo=FALSE}
options(knitr.kable.NA = '')
knitr::kable(tab[tab$beta0==0, c("rho","m","n","b0_type1","b0_hat_bias","b1_coverage","b1_hat_bias")], 
             caption = "Least Squares Estimation\nin the presence of observation correlation \n when Beta_0=0")
```  


Notice that when $\beta_0=0$ and $\rho$ is high (0.9), the probability of rejecting the (true) 
null is high (0.81, 0.52), especially when $n>m$. In other words, the larger the clusters 
--compared to the number of clusters--the greater the type 1 error probability.  
Notice, too, that bias is not a huge concern (though things are not rosey either). 
In other words, it does seem that OLS is asymptotically unbiased for both $\beta_0$ and 
not so much for $\beta_1$. However, the bias seems to be a bigger threat, again, when $n>m$.  
Things are as expected when $rho=0$: good coverage, low type 1 error, and unbiasedness for 
both $\hat{\beta}_0$ and $\hat{\beta}_1$.  

Next, see the table for an alternative model, that is when $\beta_0 \neq 0$. ($\beta_1 = 2.3$.)  

```{r, echo=FALSE}
options(knitr.kable.NA = '')
knitr::kable(tab[tab$beta0!=0, c("rho","m","n","b0_coverage","b0_hat_bias","b1_coverage","b1_hat_bias")], 
             caption = "Least Squares Estimation\nin the presence of observation correlation\nwhen Beta_0=1.9")
```  

First thing to note is the coverage of both $\hat{\beta}_1$ and $\hat{\beta}_0$. Coverage is 
95% (under $\alpha=0.05$) when $\rho=0$ but things look horrible when $\rho$ is high. As 
with the null model above, in the presence of a random intercept, this can be thought of as a 
misspecification model (or an omitted variable bias problem) and thus both 
$\beta_0$ and $\beta_1$ have poor coverage.  
Furthermore, when $n>m$ and $\rho$ is high (0.05 or 0.9), the estimates for both $\beta_0$ 
and $\beta_1$ are biased. In other words, adding more $n$ per cluster won't solve things when 
$\rho$ is high.  

I now turn to models with a binary dependent variable. Same set up, but $Y_i \in \{0,1\}$.  


## Problem 2  

Repeat the previous exercise except using a dichotomous outcome and using logistic regression.  

The simulation set-up is very similar, except that to generate the correlated $Y_i's$, I 
use the logistic function, since I'm modeling the logit of $Y_i$, not $Y_i$ per se.  

$$
Y_i = \frac{\exp\{\beta_0 + b_i + \beta_1X_i\}}{1+\exp\{\beta_0 + b_i + \beta_1X_i\}}
$$  

where $b_i$ is constant within $m$ but different across $m$, and it's (as in the linear 
example above) $\mathcal{N}(0,\theta)$ where $\theta = -(\rho \sigma^2)/(\rho - 1)$.  

I then run `glm()` with the logit link and the binomial family. Like in problem 1, I store 
coefficients, confidence intervals, and pvalues at each iteration to compute type 1 error 
(under the null model $\beta_0=0$), coverage and bias (under the alternative model 
$\beta_0 \neq 0$).  

Table 3 shows the results for the null model, that is then $\beta_0=0$ (and $\beta_1=2.3$).  

```{r, echo=FALSE}
options(knitr.kable.NA = '')
knitr::kable(tab_logit[tab_logit$beta0==0,c("rho","m","n","b0_type1","b0_hat_bias","b1_coverage","b1_hat_bias")], 
             caption = "Logistic Regression \nin the presence of observation correlation\n when Beta_0=0")
```  
Again, the probability of a type 1 error is around $\alpha$ (i.e. 0.05) when $\rho = 0$. Things 
get dicey when $\rho$ gets high. It reaches 73% when $\rho = 0.9$. Meanwhile, the bias of 
$\hat{\beta}_1$ is also high when $\rho$ gets high, more so than in the non-binary case. So, 
regardless of the number of clusters relative to samples per cluster, things are biased 
and coverage is trash for $\hat{\beta}_1$.  
Bottom line, if things were bad in linear regression, things are *really* bad for logistic 
regression.  



```{r, echo=FALSE}
options(knitr.kable.NA = '')
knitr::kable(tab_logit[tab_logit$beta0==0,c("rho","m","n","b0_coverage","b0_hat_bias","b1_coverage","b1_hat_bias")], 
             caption = "Logistic Regression \nin the presence of observation correlation\n when Beta_0=1.9")
```  

Coverage for $\beta_0$ and $\beta_1$ is spot on when $\rho = 0$. Bias is almost nonexistent. 
But again, things go downhill as $\rho$ goes up. Coverage for both $\beta_0$ and $\beta_1$ 
is awful as $\rho$ goes up. Bias for $\hat{\beta}_1$ is also bad (thought not so for $\hat{\beta}_0$) 
as $\rho$ goes up.  

So biasedness because a worse issue in the logistic case. Adding observations within $m$,
or even $m$ did not make the problem go away.  
 

## Problem 3 
The problem with dealing with multivariate data is that there are multiple outcomes. If 
there were only one outcome variable per subject, then we can just directly use the methods 
from 570. One easy way to make the data univariate is to simply take a weighted (either equal 
or unequal weights) average of the outcome measures. In many fields and application areas, 
this is what analysts (including some statisticians) do. For now, let’s just assume that we 
use flat weights. Write a coherent discussion of the pros and cons of this mode of analysis 
from the perspective of practicality, validity, and power. Consider what situations it would 
be appropriate to use this type of analysis and under what situations such an analysis might 
even be nearly optimal. Justify your assertions and explain how you come to your conclusions.  



The major downside is that a seemingly large dataset (n=1,000) but with only 5 individuals 
(meaning each individual has 20 measurements) taking the mean within individual will reduce 
the sample size quite a bit. In other words, we significant reduce power by doing this.  

Furthermore, for this exercise we've been assuming a balanced panel, that is, a situation 
where each cluster or individual has the same number of measurements. This does not have 
to be true necessarily. Taking the mean across many observations within a cluster may be fine, 
but when a cluster only has a handful of observation, the idea is not that appealing anymore.  



\newpage

## Apendix  

Code  
```
set.seed(571)
rm(list = ls())
alpha <- 0.05
N <- 1e3

# ~~~~~~~~~ #
# Problem 1 #
# ~~~~~~~~~ #

simul_reg <- function(m, n, rho, beta0, beta1, sigma=1, alpha=alpha) {
  
  # create random intercept b_rand ~ N(0,theta)
  theta = -(rho*sigma^2)/(rho-1)
  b_r <- rep(rnorm(m, 0, theta),n)
  
  # Create e vector
  e <- rnorm(m*n, 0, sigma)
  
  # Create id vector for sanity checks
  id <- rep(1:m,n)
  
  # Create X vector where trt is randomly assigned per id
  trt <- sample(c(0,1),1)
  X <- rep(trt, m*n)
  if(trt==1) { 
    X[id %% 2 == 0] <- 0 
  }else { 
    X[id %% 2 == 0] <- 1 
  }
  
  # Create Y vector with true relationship
  Y <- beta0 + beta1*X + b_r + e 
  
  data.frame(id=id, Y=Y, X=X, b_r=b_r, e=e) |> dplyr::arrange(id)
  
  # Fit linear model
  fit <- lm(Y~X)
  beta_hat <- coef(fit)
  beta_hat_SE <- sqrt(diag(vcov(fit)))
  CI_beta_hat <- matrix(c(beta_hat - abs(qt(alpha/2, m*n-2))*beta_hat_SE, 
                          beta_hat + abs(qt(alpha/2, m*n-2))*beta_hat_SE), ncol=2, nrow=2)
  #CI_beta_hat <- confint(fit)
  inside <- c(as.numeric(CI_beta_hat[1,1] <= beta0 & beta0 <= CI_beta_hat[1,2]),
              as.numeric(CI_beta_hat[2,1] <= beta1 & beta1 <= CI_beta_hat[2,2]))
  pval <- summary(fit)$coefficients[,4]
  
  #type 1 error: rejecting the null when the null is true
  type_1 <- c(NA,NA)
  if(beta0==0) {
    type_1[1] <- as.numeric(pval[1]<0.05)
  }
  if(beta1==0) {
    type_1[2] <- as.numeric(pval[2]<0.05)
  }
  
  o <- list("b0_inside" = inside[[1]], 
            "b0_type_1" = type_1[[1]], 
            "b0_hat"    = beta_hat[[1]],
            "b1_inside" = inside[[2]], 
            "b1_type_1" = type_1[[2]], 
            "b1_hat"    = beta_hat[[2]])
  return(o)
}

#~~~~~~~~~~~#
# Problem 2 #
#~~~~~~~~~~~#
simul_logit <- function(m, n, rho, beta0=1, beta1=0, alpha=0.05, sigma=1) {
  
  # create correlated X which will determine probabilities p
  theta = -(rho*sigma^2)/(rho-1)
  b_r <- rep(rnorm(m, 0, theta),n)
  X <- rnorm(n*m, 0,1)
  p <- exp(beta0+b_r+beta1*X)/(1+exp(beta0+b_r+beta1*X))
  Y <- rbinom(n=m*n, size=1, prob=p)
  id <- rep(1:m, n)
  df <- data.frame(id=id, Y=Y, 
                   X=matrix(X, nrow=m*n, ncol=1), 
                   p=matrix(p, nrow=m*n, ncol=1)) |> dplyr::arrange(id)
  df

  # fit logistic model
  cz <- qnorm(1-alpha/2, 0, 1)
  fit <- summary(glm(Y~X, data=df, family=binomial(link="logit")))$coefficients
  beta_hat  <- fit[,1]
  inside    <- c(as.numeric(fit[1,1]-cz*fit[1,2] < beta0 & beta0 < fit[1,1]+cz*fit[1,2]),
                 as.numeric(fit[2,1]-cz*fit[2,2] < beta1 & beta1 < fit[2,1]+cz*fit[2,2]))
  pval <- fit[,4]
  
  #type 1 error: rejecting the null when the null is true
  type_1 <- c(NA,NA)
  if(beta0==0) {
    type_1[1] <- as.numeric(pval[1]<0.05)
  }
  if(beta1==0) {
    type_1[2] <- as.numeric(pval[2]<0.05)
  }
  
  # output per iteration
  o <- list("b0_inside" = inside[[1]], 
            "b0_type_1" = type_1[[1]], 
            "b0_hat"    = beta_hat[[1]],
            "b1_inside" = inside[[2]], 
            "b1_type_1" = type_1[[2]], 
            "b1_hat"    = beta_hat[[2]])
  return(o)
}



# ~~~~~~~~~~~~~~~~~~~~~~~ #
# create table of results #
# ~~~~~~~~~~~~~~~~~~~~~~~ #

run_sims <- function(N, m, n, rho, beta0, beta1, alpha=0.05, sigma=1, flavor="ols") {
  coverage_b0 <- rep(0,N)    ;coverage_b1 <- rep(0,N)
  type_1_b0   <- rep(0,N)    ;type_1_b1   <- rep(0,N)
  b0_hat      <- rep(0,N)    ;b1_hat      <- rep(0,N)
  for(i in 1:N) {
    if(flavor=="logit") {
      r <- simul_logit(m=m, n=n, rho=rho, beta1=beta1, beta0=beta0, sigma=sigma)
    } else{
      r <-   simul_reg(m=m, n=n, rho=rho, beta0=beta0, beta1=beta1, sigma=sigma)
    }
    coverage_b0[i] <- r[[1]] 
    type_1_b0[i]   <- r[[2]] 
    b0_hat[i]      <- r[[3]]
    coverage_b1[i] <- r[[4]] 
    type_1_b1[i]   <- r[[5]] 
    b1_hat[i]      <- r[[6]]    
  }
  o <- list("coverage_b0" = mean(coverage_b0) ,"coverage_b1" = mean(coverage_b1),
            "type_1_b0"   = mean(type_1_b0)   ,"type_1_b1"   = mean(type_1_b1),
            "b0_hat"      = beta0-mean(b0_hat),"b1_hat"      = beta1-mean(b1_hat))
  return(o)
}

tab <- dplyr::tibble(beta0=numeric(), beta1=numeric(),
                     rho=numeric(), m=numeric(), n=numeric(), 
                     b0_coverage=numeric(), 
                     b0_type1=numeric(), 
                     b0_hat_bias=numeric(),
                     b1_coverage=numeric(), 
                     b1_type1=numeric(), 
                     b1_hat_bias=numeric())

tab_logit <- dplyr::tibble(beta0=numeric(), beta1=numeric(),
                     rho=numeric(), m=numeric(), n=numeric(), 
                     b0_coverage=numeric(), 
                     b0_type1=numeric(), 
                     b0_hat_bias=numeric(),
                     b1_coverage=numeric(), 
                     b1_type1=numeric(), 
                     b1_hat_bias=numeric())

for(beta0 in c(0,1.9)) {
  for(beta1 in c(2.3)) {
    for(rho in c(0.9, 0.5, 0)) {
      print(paste0("beta0=",beta0," beta1=",beta1, " rho=", rho))
      for(m in c(10,50)) {
        for(n in c(50,10)) {
          if(m!=n) {  
            r <- run_sims(N=N, m=m, n=n, rho=rho, beta1=beta1, beta0=beta0, flavor="ols")
            tab <- dplyr::add_row(tab, 
                                  "beta0"=beta0, "beta1"=beta1, 
                                  "rho"=rho, "m"=m, "n"=n, 
                                  "b0_coverage"=r[[1]], 
                                  "b0_type1"=r[[3]], 
                                  "b0_hat_bias"=round(r[[5]],2),
                                  "b1_coverage"=r[[2]], 
                                  "b1_type1"=r[[4]], 
                                  "b1_hat_bias"=round(r[[6]],2))
            r <- run_sims(N=N, m=m, n=n, rho=rho, beta1=beta1, beta0=beta0, flavor="logit")
            tab_logit <- dplyr::add_row(tab_logit, 
                                  "beta0"=beta0, "beta1"=beta1, 
                                  "rho"=rho, "m"=m, "n"=n, 
                                  "b0_coverage"=r[[1]], 
                                  "b0_type1"=r[[3]], 
                                  "b0_hat_bias"=round(r[[5]],2),
                                  "b1_coverage"=r[[2]], 
                                  "b1_type1"=r[[4]], 
                                  "b1_hat_bias"=round(r[[6]],2))            
            
          }
        }
      }
    }
  }
}
tab
tab_logit
objs <- ls()
objs <- ls()[!(ls() %in% c("tab","tab_logit","simul_logit","simul_reg","run_sims"))]
rm(list = objs)
```

























